---
title: "Dog Vocalizations"
author: "Kaili Saffran"
date: "2/3/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(av)
require(data.table)
```


```{r}
# # Accessing data no matter what working directory you are in:
# library(here)
# set_here()
# # enclose data file in here()
# # create project in Github
```


# Data
## Emotions/Characteristics expressed by dogs:
#### Anxiety
```{r}
# get information on each WAV file:
av_media_info("Anxiety_1.wav")
# convert WAV files to have only one channel:
data_anx1 <- read_audio_bin("Anxiety_1.wav", channels = 1)
# plot a frequency plot of the WAV file to see its structure:
plot(data_anx1, type = 'l')

# -----------------------------------------------------
# not needed, but cool to look at:
# # changes WAV file to frequency domain to get a spectrogram:
# dataf_anx1 <- read_audio_fft("Anxiety_1.wav")
# dim(dataf_anx1)
# plot(dataf_anx1)
# ------------------------------------------------------
```


```{r}
av_media_info("Anxiety_2.wav")
data_anx2 <- read_audio_bin("Anxiety_2.wav", channels = 1)
plot(data_anx2, type = 'l')
```

```{r}
# # convert mp3 file to WAV file:
# av_media_info("/Users/kailisaffran/Documents/Dog Vocalizations/Anxiety3.mp3")
# av_audio_convert(audio = "/Users/kailisaffran/Documents/Dog Vocalizations/Anxiety3.mp3", output = "Anxiety3.wav", total_time = 12)
av_media_info("Anxiety3.wav")
data_anx3 <- read_audio_bin("Anxiety3.wav", channels = 1)
plot(data_anx3, type = 'l')
```

```{r}
# av_media_info("/Users/kailisaffran/Documents/Dog Vocalizations/Anxiety4.mp3")
# av_audio_convert(audio = "/Users/kailisaffran/Documents/Dog Vocalizations/Anxiety4.mp3", output = "Anxiety4.wav")
av_media_info("Anxiety4.wav")
data_anx4 <- read_audio_bin("Anxiety4.wav", channels = 1)
plot(data_anx4, type = 'l')
```

#### Excitement
```{r}
av_media_info("Excite_4.wav")
data_exc4 <- read_audio_bin("Excite_4.wav", channels = 1)
plot(data_exc4, type = 'l')
```


```{r}
av_media_info("Excite_6.wav")
data_exc6 <- read_audio_bin("Excite_6.wav", channels = 1)
plot(data_exc6, type = 'l')
```


```{r}
# av_media_info("/Users/kailisaffran/Documents/Dog Vocalizations/Excite7.mp3")
# av_audio_convert(audio = "/Users/kailisaffran/Documents/Dog Vocalizations/Excite7.mp3", output = "Excite7.wav", start_time = 0, total_time = 5)
av_media_info("Excite7.wav")
data_exc7 <- read_audio_bin("Excite7.wav", channels = 1)
plot(data_exc7, type = 'l')
```


```{r}
av_media_info("Excite_10.wav")
data_exc10 <- read_audio_bin("Excite_10.wav", channels = 1)
plot(data_exc10, type = 'l')
```

#### Playful
```{r}
# av_media_info("/Users/kailisaffran/Documents/Dog Vocalizations/Play3.mp3")
# av_audio_convert(audio = "/Users/kailisaffran/Documents/Dog Vocalizations/Play3.mp3", output = "Play3.wav")
av_media_info("Play3.wav")
data_play3 <- read_audio_bin("Play3.wav", channels = 1)
plot(data_play3, type = 'l')
```


```{r}
av_media_info("Play_5.wav")
data_play5 <- read_audio_bin("Play_5.wav", channels = 1)
plot(data_play5, type = 'l')
```

```{r}
# av_media_info("/Users/kailisaffran/Documents/Dog Vocalizations/Play6.mp3")
# av_audio_convert(audio = "/Users/kailisaffran/Documents/Dog Vocalizations/Play6.mp3", output = "Play6.wav", start_time = 0, total_time = 6)
av_media_info("Play6.wav")
data_play6 <- read_audio_bin("Play6.wav", channels = 1)
plot(data_play6, type = 'l')
```

```{r}
av_media_info("Play_1.wav")
data_play1 <- read_audio_bin("Play_1.wav", channels = 1)
plot(data_play1, type = 'l')
```

#### Protection
```{r}
# av_media_info("/Users/kailisaffran/Documents/Dog Vocalizations/Protect1.mp3")
# av_audio_convert(audio = "/Users/kailisaffran/Documents/Dog Vocalizations/Protect1.mp3", output = "Protect1.wav", start_time = 5, total_time = 20)
av_media_info("Protect1.wav")
data_protect1 <- read_audio_bin("Protect1.wav", channels = 1)
plot(data_protect1, type = 'l')
```


```{r}
# av_media_info("/Users/kailisaffran/Documents/Dog Vocalizations/Protect3.mp3")
# av_audio_convert(audio = "/Users/kailisaffran/Documents/Dog Vocalizations/Protect3.mp3", output = "Protect3.wav", start_time = 0, total_time = 11)
av_media_info("Protect3.wav")
data_protect3 <- read_audio_bin("Protect3.wav")
plot(data_protect3, type = 'l')
```

```{r}
# av_media_info("/Users/kailisaffran/Documents/Dog Vocalizations/Protect4.mp3")
# av_audio_convert(audio = "/Users/kailisaffran/Documents/Dog Vocalizations/Protect4.mp3", output = "Protect4.wav")
av_media_info("Protect4.wav")
data_protect4 <- read_audio_bin("Protect4.wav", channels = 1)
plot(data_protect4, type = 'l')
```


```{r}
av_media_info("Protect_5.wav")
data_protect5 <- read_audio_bin("Protect_5.wav")
plot(data_protect5, type = 'l')
```

#### Upset/Wants Something
```{r}
# av_media_info("/Users/kailisaffran/Documents/Dog Vocalizations/Upset1.mp3")
# av_audio_convert(audio = "/Users/kailisaffran/Documents/Dog Vocalizations/Upset1.mp3", output = "Upset1.wav")
av_media_info("Upset1.wav")
data_upset1 <- read_audio_bin("Upset1.wav", channels = 1)
plot(data_upset1, type = 'l')
```

```{r}
# av_media_info("/Users/kailisaffran/Documents/Dog Vocalizations/Upset4.mp3")
# av_audio_convert(audio = "/Users/kailisaffran/Documents/Dog Vocalizations/Upset4.mp3", output = "Upset4.wav")
av_media_info("Upset4.wav")
data_upset4 <- read_audio_bin("Upset4.wav", channels = 1)
plot(data_upset4, type = 'l')
```

```{r}
# av_media_info("/Users/kailisaffran/Documents/Dog Vocalizations/Upset6.mp3")
# av_audio_convert(audio = "/Users/kailisaffran/Documents/Dog Vocalizations/Upset6.mp3", output = "Upset6.wav", total_time = 5)
av_media_info("Upset6.wav")
data_upset6 <- read_audio_bin("Upset6.wav")
plot(data_upset6, type = 'l')
```


```{r}
av_media_info("Upset_3.wav")
data_upset3 <- read_audio_bin("Upset_3.wav", channels = 1)
plot(data_upset3, type = 'l')
```

# Dynamic Time Warping
```{r}
# # this function does not work do to the size of the WAV files:
# require(dtw)
# dtwDist

# create zero vectors to get all WAV files to be of the same length:
# largest WAV file is 594432 elements after referencing the environment:
a <- 594432-length(data_anx1)
a2 <- (rep(0, a))
b <- 594432-length(data_anx2)
b2 <- (rep(0, b))
c <- 594432-length(data_anx3)
c2 <- (rep(0, c))
d <- 594432-length(data_anx4)
d2 <- (rep(0, d))
e <- 594432-length(data_upset1)
e2 <- (rep(0, e))
f <- 594432-length(data_upset3)
f2 <- (rep(0, f))
g <- 594432-length(data_upset4)
g2 <- (rep(0, g))
h <- 594432-length(data_upset6)
h2 <- (rep(0, h))
i <- 594432-length(data_exc10)
i2 <- (rep(0, i))
j <- 594432-length(data_exc4)
j2 <- (rep(0, j))
k <- 594432-length(data_exc6)
k2 <- (rep(0, k))
l <- 594432-length(data_exc7)
l2 <- (rep(0, l))
m <- 594432-length(data_play1)
m2 <- (rep(0, m))
n <- 594432-length(data_play3)
n2 <- (rep(0, n))
o <- 594432-length(data_play5)
o2 <- (rep(0, o))
p <- 594432-length(data_play6)
p2 <- (rep(0, p))
q <- 594432-length(data_protect1)
q2 <- (rep(0, q))
r <- 594432-length(data_protect3)
r2 <- (rep(0, r))
s <- 594432-length(data_protect4)
s2 <- (rep(0, s))
t <- 594432-length(data_protect5)
t2 <- (rep(0, t))

# ------------------------------------------------------------
# not needed, convert WAV files to a time series:
# ts(data = c(a,b,c,d,e,f,g,h,i,j,k,l,m,n,o,p,q,r,s,t), class = "mts")
# ------------------------------------------------------------
```


```{r}
# --------------------------------------------------------------
# not needed:
# create zero matrix with rows representative of each time series and columns the length of largest WAV file:
# Bark_Matrix <- matrix(0, nrow = 20, ncol = 594432)
# --------------------------------------------------------------

# add vectors of 0 to end of each WAV file so they all have equal lengths:
a3 <- append(as.vector(data_anx1), a2, after = length(data_anx1))
b3 <- append(as.vector(data_anx2), b2, after = length(data_anx2))
c3 <- append(as.vector(data_anx3), c2, after = length(data_anx3))
d3 <- append(as.vector(data_anx4), d2, after = length(data_anx4))
e3 <- append(as.vector(data_upset1), e2, after = length(data_upset1))
f3 <- append(as.vector(data_upset3), f2, after = length(data_upset3))
g3 <- append(as.vector(data_upset4), g2, after = length(data_upset4))
h3 <- append(as.vector(data_upset6), h2, after = length(data_upset6))
i3 <- append(as.vector(data_exc10), i2, after = length(data_exc10))
j3 <- append(as.vector(data_exc4), j2, after = length(data_exc4))
k3 <- append(as.vector(data_exc6), k2, after = length(data_exc6))
l3 <- append(as.vector(data_exc7), l2, after = length(data_exc7))
m3 <- append(as.vector(data_play1), m2, after = length(data_play1))
n3 <- append(as.vector(data_play3), n2, after = length(data_play3))
o3 <- append(as.vector(data_play5), o2, after = length(data_play5))
p3 <- append(as.vector(data_play6), p2, after = length(data_play6))
q3 <- append(as.vector(data_protect1), q2, after = length(data_protect1))
r3 <- append(as.vector(data_protect3), r2, after = length(data_protect3))
s3 <- append(as.vector(data_protect4), s2, after = length(data_protect4))
t3 <- append(as.vector(data_protect5), t2, after = length(data_protect5))
```


```{r}
# ------------------------------------------------
# # not needed:
# # used as a reference:
# # put rows of new time series into the matrix:
# Bark_Matrix[1,]  <- a3
# Bark_Matrix[2,]  <- b3
# Bark_Matrix[3,]  <- c3
# Bark_Matrix[4,]  <- d3
# Bark_Matrix[5,]  <- e3
# Bark_Matrix[6,]  <- f3
# Bark_Matrix[7,]  <- g3
# Bark_Matrix[8,]  <- h3
# Bark_Matrix[9,]  <- i3
# Bark_Matrix[10,]  <- j3
# Bark_Matrix[11,]  <- k3
# Bark_Matrix[12,]  <- l3
# Bark_Matrix[13,]  <- m3
# Bark_Matrix[14,]  <- n3
# Bark_Matrix[15,]  <- o3
# Bark_Matrix[16,]  <- p3
# Bark_Matrix[17,]  <- q3
# Bark_Matrix[18,]  <- r3
# Bark_Matrix[19,]  <- s3
# Bark_Matrix[20,]  <- t3
# ----------------------------------------------

# create dissimilarity matrix: need to use a different method for dynamic time warping since computer does not have enough memory to run with "dist" function:
  # Bark_Dist <- dist(Bark_Matrix, method = "DTW")
```


# Manually performing DTW:
## Options: Simulated Annealing, "L"/Stair, or Random Sampling Method:
### staircase method:
### this approach will not find the global minimum:
### D(i, j) = d(i, j) + min(D(i − 1, j − 1), D(i, j − 1), D(i − 1, j))

# Random Sampling Method: 
## Used this method because it made the most sense and was most efficient of the methods I discussed and developed
```{r Sample 4x4 DTW Matrix}
# ---------------------------------------------------------------------------
# do not need to run now that we know it works:
# this is for a smaller matrix to get a path simulation working:
# use as a reference for real simulation:
set.seed(1025)
sample_bark <- matrix(data = sample(c(1:10), 16, replace = TRUE), nrow = 4, ncol = 4, byrow = T)

# Conditions:
# sample different routes of paths ~ law of large numbers
# look at distributions and compare
# store the smallest sum
# probabilities of going in one direction: 1/3

# bottom left-hand corner is the starting point:

R <- 100
df_sums <- data.table("Sums" = rep(0, R))

for (r in 1:R) {

#create empty vector to store the path for each simulation:
path_bark <- c()
i <- nrow(sample_bark)
j <- 1
n <- 1
start <- sample_bark[i, j]
path_bark[n] <- start

while (i != 1 | j != ncol(sample_bark)) {
  if (i != 1 & j != ncol(sample_bark)) {
    path_1 <- sample_bark[(i - 1), j]
    path_2 <- sample_bark[i, (j + 1)]
    path_3 <- sample_bark[(i - 1), (j + 1)]
    paths <- c(path_1, path_2, path_3)
    next.place <- sample(paths, 1)
    if (next.place == path_1) {
      i <- i - 1
    } else if (next.place == path_2) {
      j <- j + 1
    } else if (next.place == path_3) {
      i <- i - 1
      j <- j + 1
    }
    path_bark[n + 1] <- next.place
    n <- n + 1
  }
  if (i == 1) {
    if (j < ncol(sample_bark)) {
    next.place <- sum(sample_bark[i, (j + 1):ncol(sample_bark)])
    path_bark[n + 1] <- next.place
    j <- ncol(sample_bark)
    }
  } else if (j == ncol(sample_bark)) {
    next.place <- sum(sample_bark[1:(i - 1), j])
    path_bark[n + 1] <- next.place
    i <- 1
  }
  store_sum <- sum(path_bark)
}
df_sums$Sums[r] <- store_sum
}
store_min <- min(df_sums$Sums)
# --------------------------------------------------------------------------
```


```{r Finding difference between two time series}
# --------------------------------------------------------------------
# not needed: use as reference for real simulation:
# first did calculations by hand and then compared to code to make sure it worked:

for (row1 in 1:8) {
   for (row2 in 1:8) {
     Euc_Dist <- Bark_Matrix[1, row1] - Bark_Matrix[2, row2]
     Result <- Euc_Dist
    print(Result)
   }
}
# --------------------------------------------------------------------
```


```{r Creating dissimilarity matrix}

set.seed(1025)

# want to generate matrix as we sample a path:
# create dissimilarity matrix to store min between times series in:
Dis_Matrix <- matrix(data = 0, nrow = 20, ncol = 20, byrow = T,
                     dimnames = list(c("a3","b3","c3","d3","e3","f3","g3","h3","i3","j3","k3","l3","m3","n3",
                                       "o3","p3","q3","r3","s3","t3"),
                                     c("a3","b3","c3","d3","e3","f3","g3","h3","i3","j3","k3","l3","m3","n3",
                                       "o3","p3","q3","r3","s3","t3")))

# create a list of the time series:
Time_Series <- c(a3,b3,c3,d3,e3,f3,g3,h3,i3,j3,k3,l3,m3,n3,o3,p3,q3,r3,s3,t3)
col_place <- 0
row_place <- 1
x <- 0

for (ts1 in 0:18) {
  col_place <- col_place + 1
  x <- x + 1
  for (ts2 in x:19) {
    row_place <- row_place + 1

R <- 1
df_sums <- data.table("Sums" = rep(0, R))

for (r in 1:R) {

path_bark <- c()
i <- ncol(Bark_Matrix)
j <- 1
n <- 1
start <- Time_Series[j+(ts1*ncol(Bark_Matrix))] - Time_Series[i+(ts2*ncol(Bark_Matrix))]
path_bark[n] <- start

while (i != 1 | j != ncol(Bark_Matrix)) {
  if (i != 1 & j != ncol(Bark_Matrix)) {
    # replace following code by putting it in the if statements:
    # path_1 <- Time_Series[j+(ts1*ncol(Bark_Matrix))] - Time_Series[(i-1)+(ts2*ncol(Bark_Matrix))]
    # path_2 <- Time_Series[(j+1)+(ts1*ncol(Bark_Matrix))] - Time_Series[i+(ts2*ncol(Bark_Matrix))]
    # path_3 <- Time_Series[(j+1)+(ts1*ncol(Bark_Matrix))] - Time_Series[(i-1)+(ts2*ncol(Bark_Matrix))]
    # paths <- c(path_1, path_2, path_3)
    next.place <- sample(c("path_1","path_2","path_3"), 1)
    if (next.place == "path_1") {
      current.place <- Time_Series[j+(ts1*ncol(Bark_Matrix))] - Time_Series[(i-1)+(ts2*ncol(Bark_Matrix))]
      i <- i - 1
      # breaking apart this code piece by piece, for the first step, path_2 and path_3 are equal...problematic:
      # need to add/fix code so if two paths are equal, it only chooses one:
    } else if (next.place == "path_2") {
      current.place <- Time_Series[(j+1)+(ts1*ncol(Bark_Matrix))] - Time_Series[i+(ts2*ncol(Bark_Matrix))]
      j <- j + 1
    } else if (next.place == "path_3") {
      current.place <- Time_Series[(j+1)+(ts1*ncol(Bark_Matrix))] - Time_Series[(i-1)+(ts2*ncol(Bark_Matrix))]
      i <- i - 1
      j <- j + 1
    }
    path_bark[n + 1] <- current.place
    n <- n + 1
  }
  if (i == 1) {
    if (j < ncol(Bark_Matrix)) {
    current.place <- sum(Time_Series[(j+1)+(ts1*ncol(Bark_Matrix)):ncol(Bark_Matrix)+(ts1*ncol(Bark_Matrix))]
                      - Time_Series[i+(ts2*ncol(Bark_Matrix))])
    path_bark[n + 1] <- current.place
    j <- ncol(Bark_Matrix)
    }
  } else if (j == ncol(Bark_Matrix)) {
    current.place <- sum(Time_Series[j+(ts1*ncol(Bark_Matrix))] - 
                      Time_Series[1+(ts2*ncol(Bark_Matrix)):(i-1)+(ts2*ncol(Bark_Matrix))])
    path_bark[n + 1] <- current.place
    i <- 1
  }
  store_sum <- sum(path_bark)
}
df_sums$Sums[r] <- store_sum
}
store_min <- min(df_sums$Sums)
Dis_Matrix[row_place, col_place] <- store_min
  }
}

# save(Dis_Matrix, file = "Dis_Matrix.Rdata")
# after an hour, ts2 was on 9 and ts1 on 0
# need a more efficient way
# this way takes a long time because the loop has to search through the entire vector to compare two time series
```



```{r Comparing two time series at a time with a LIST}

# create dissimilarity matrix to store min between times series in:
Dis_Matrix <- matrix(data = 0, nrow = 20, ncol = 20, byrow = T,
                     dimnames = list(c("a3","b3","c3","d3","e3","f3","g3","h3","i3","j3","k3","l3","m3","n3",
                                       "o3","p3","q3","r3","s3","t3"),
                                     c("a3","b3","c3","d3","e3","f3","g3","h3","i3","j3","k3","l3","m3","n3",
                                       "o3","p3","q3","r3","s3","t3")))

# create a list of the time series:
  # Time_Series <- list(a3,b3,c3,d3,e3,f3,g3,h3,i3,j3,k3,l3,m3,n3,o3,p3,q3,r3,s3,t3)
  # save(Time_Series, file = "Time_Series_Data.Rdata")
# set column place for filling in the dissimilarity matrix:
col_place <- 0
# set row place for filling in the dissimilarity matrix:
row_place <- 1
x <- 1
# time series length:
TSL <- ncol(Bark_Matrix)

for (ts1 in 1:19) {
  col_place <- col_place + 1
  x <- x + 1
  for (ts2 in x:20) {
    row_place <- row_place + 1

# Number of simulations:
R <- 1
# store the sums of each path:
df_sums <- data.table("Sums" = rep(0, R))

for (r in 1:R) {

path_bark <- c()
i <- TSL
j <- 1
n <- 1
start <- Time_Series[[ts1]][1] - Time_Series[[ts2]][TSL]
path_bark[n] <- start

while (i != 1 | j != TSL) {
  if (i != 1 & j != TSL) {

    next.place <- sample(c("path_1","path_2","path_3"), 1)
    if (next.place == "path_1") {
      current.place <- Time_Series[[ts1]][j] - Time_Series[[ts2]][(i-1)]
      i <- i - 1
    } else if (next.place == "path_2") {
      current.place <- Time_Series[[ts1]][(j+1)] - Time_Series[[ts2]][i]
      j <- j + 1
    } else if (next.place == "path_3") {
      current.place <- Time_Series[[ts1]][(j+1)] - Time_Series[[ts2]][(i-1)]
      i <- i - 1
      j <- j + 1
    }
    path_bark[n + 1] <- current.place
    n <- n + 1
  }
  if (i == 1) {
    if (j < TSL) {
    current.place <- sum(Time_Series[[ts1]][(j+1):TSL]
                      - Time_Series[[ts2]][i])
    path_bark[n + 1] <- current.place
    j <- TSL
    }
  } else if (j == TSL) {
    current.place <- sum(Time_Series[[ts1]][j] - 
                      Time_Series[[ts2]][1:(i-1)])
    path_bark[n + 1] <- current.place
    i <- 1
  }
  store_sum <- sum(path_bark)
}
df_sums$Sums[r] <- store_sum
}
store_min <- min(df_sums$Sums)
Dis_Matrix[row_place, col_place] <- store_min
  }
}


# save(Dis_Matrix, file = "Dis_Matrix.Rdata")
# after an hour, ts2 was on 9 and ts1 on 0 --> no improvement
# still takes a while to run
```


```{r Most efficient way to do it}
set.seed(1025)

x <- 1
difference <- c()
n <- 0
for (ts_1 in 1:19) {
  x <- x + 1
  for (ts_2 in x:20) {
    for (i in 1:10000) {
      length(Time_Series[[1]]) -> TSL
      
      sample(
        x = c(0, 1),
        size = 2 * TSL,
        # This is the maximum number needed
        prob = c(1 / 3, 2 / 3),
        replace = TRUE
      ) -> x_s
      
      sample(
        x = c(0, 1),
        size = 2 * TSL,
        # This is the maximum number needed
        prob = c(1 / 3, 2 / 3),
        replace = TRUE
      ) -> y_s
    
    which(cumsum(x_s) == TSL - 1) -> x_m
    which(cumsum(y_s) == TSL - 1) -> y_m
    
    if (x_m < y_m) {
      0 -> x_s[(x_m + 1):y_m]
      x_s[1:y_m] -> x_steps
      y_s[1:y_m] -> y_steps
    } else {
      if (y_m < x_m) {
        0 -> y_s[(y_m + 1):x_m]
        y_s[1:x_m] -> y_steps
        x_s[1:x_m] -> x_steps
      } else {
        x_s[1:x_m] -> x_steps
        y_s[1:y_m] -> y_steps
      }
    }
      cumsum(x_steps) -> x_val
      cumsum(y_steps) -> y_val
      
      Time_Series[[ts_1]][1] - Time_Series[[ts_2]][1] -> difference[i+(10000*n)]
      sum(Time_Series[[ts_1]][x_val] -
            Time_Series[[ts_2]][y_val]) + difference[i] -> difference[i+(10000*n)]
    }
    n <- n + 1
  }
}
save(difference, file = "difference_data.Rdata")
```


```{r Euclidean Distance}

# this is ultimately what is being calculated above in the for loops that find the difference between points in different time series:
library(philentropy)
one <- 1:10
two <- 5:14
three <- rbind(one, two)
distance(three, method = "euclidean")

```


```{r Clustering time series}

hclust(Dis_Matrix, method = "complete", members = NULL, labels = TRUE)

```
